\begin{authorship}J\'er\'emy Barbay, inspired by Unknown\end{authorship}
\begin{usage}
A similar one was used in Assignment 2, Year 2002, Fall Term, School of Computer Science at the University of Waterloo.
Midterm, Year 2005, Winter Term, School of Computer Science at the University of Waterloo.
\end{usage}
 Consider a dictionary implemented as a linear array.
%
Let $A[0..6]$ be the dictionary, with elements $A[i]=i$ for $0 \leq i
\leq 6$.
%
You have seen the Move-To-Front (MTF) and Transpose heuristics in
class, but there is another possible heuristic: Move-To-Middle (MTM).
%
The MTM heuristic is like the MTF heuristic, except that once
an element is found at position $p$, it is moved halfway to the front,
to the ``middle'' position $\lfloor p/2\rfloor$.

\begin{enumerate}

\item For each of MTF, Transpose, MTM, show the array after searching for
the following keys in the given order: $3; 1; 6; 6.$
%
For each heuristic, compute the total number of (key) comparisons
required to execute all of these searches.

\begin{spaceForAnswer}
\begin{tabular}{|c|p{3.2cm}|p{3.2cm}|p{3.2cm}|}
\hline
Search & MTF & Transpose & MTM\\
\hline
\hline
Start&
\multicolumn{3}{|c|}{\tt [0,1,2,3,4,5,6]}\\
\hline
\hline
3
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
\\
&
c= \hspace{.5cm} &
c= \hspace{.5cm} &
c= \hspace{.5cm} \\
\hline
1
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
\\
&
c= \hspace{.5cm} &
c= \hspace{.5cm} &
c= \hspace{.5cm} \\
\hline
6
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
\\
&
c= \hspace{.5cm} &
c= \hspace{.5cm} &
c= \hspace{.5cm} \\
\hline
6
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
&$\begin{array}{|c|c|c|c|c|c|c|}\hline\ &\ &\ &\ &\ &\ &\ \\ \hline\end{array}$
\\
&
c= \hspace{.5cm} &
c= \hspace{.5cm} &
c= \hspace{.5cm} \\
\hline
\hline
Total comparisons&
  &  &  \\
\hline
\end{tabular}
\end{spaceForAnswer}
\begin{solution}
\begin{tabular}{|c|l|l|l|}
\hline
Search & MTF & Transpose & MTM\\
\hline
\hline
Start&
\multicolumn{3}{|c|}{\tt [0,1,2,3,4,5,6]}\\
\hline
\hline
3&
{\tt [3,0,1,2,4,5,6]}&
{\tt [0,1,3,2,4,5,6]}&
{\tt [0,3,1,2,4,5,6]}\\
&
c=4&
c=4&
c=4\\
\hline
1&
{\tt [1,3,0,2,4,5,6]}&
{\tt [1,0,3,2,4,5,6]}&
{\tt [0,1,3,2,4,5,6]}\\
&
c=3&
c=2&
c=3\\
\hline
6&
{\tt [6,1,3,0,2,4,5]}&
{\tt [1,0,3,2,4,6,5]}&
{\tt [0,1,3,6,2,4,5]}\\
&
c=7&
c=7&
c=7\\
\hline
6&
{\tt [6,1,3,0,2,4,5]}&
{\tt [1,0,3,2,6,4,5]}&
{\tt [0,6,1,3,2,4,5]}\\
&
c=1&
c=5&
c=4\\
\hline
\hline
Total comparisons&
15&19&18\\
\hline
\end{tabular}

\end{solution}

\item Given the following distribution of search queries on the keys
of the dictionary, what is the expected number of comparisons needed to
search in an optimal static ordering of the list?
%
$$
\Pr[0]{=}.01;\
\Pr[1]{=}.3;\
\Pr[2]{=}.09;\
\Pr[3]{=}.15;\
\Pr[4]{=}.2;\
\Pr[5]{=}.15;\
\Pr[6]{=}.1
$$
Give the optimal static ordering, the general formula to compute the
expected number of comparisons, and the details of your calculations.

\begin{spaceForAnswer}

\vfil

\end{spaceForAnswer}
%
\begin{solution}
One optimal static ordering is $1,4,5,3,6,2,0$ (the other one is $1,4,3,5,6,2,0$).

With the following notation, where $p_i$ corresponds to the $i$th
probability in decreasing order starting from $p_1$, the average
complexity of a search in this list is then $\sum_{i=1}^{7} i p_i$:
$$
\begin{array}{|c|c|c|c|c|c|c|c|} \hline
i   & 1  & 2  & 3 & 4 & 5 & 6 & 7 \\ \hline
p_i & .3 & .2 &.15&.15&.1 &.09&.01\\  \hline
\end{array}
$$

Hence the average complexity is
$$
\begin{array}{llllllll}
&   1 * .3
& + 2 * .2
& + 3 * .15
& + 4 * .15
& + 5 * .1
& + 6 * .09
& + 7 * .01
\\
=
&   .3
& + .4
& + .45
& + .6
& + .5
& + .54
& + .07
\\
=
& 2.86
\end{array}
$$
\end{solution}
\begin{markingScheme}
$2$ points for the optimal ordering,
$2$ points for the formula,
$1$ point for the correct computation,
for a total of $5$ points.
\end{markingScheme}
\end{enumerate}
